\plaintitle

### 2020-05-13 paper/paper.tex
little tweak to paper
finished paper

% Paper metadata (use plain text, for PDF inclusion and later
% re-using, if desired).  Use \emtpyauthor when submitting for review
% so you remain anonymous.
\def\plaintitle{Abstract Program Visualization for Model-View-Update
User Interfaces}
\def\plainauthor{Geoffrey Litt}
\def\emptyauthor{}
\def\plainkeywords{Program visualization, program understanding, debugging}
\def\plaingeneralterms{Program visualization, program understanding, debugging}

@@

\begin{abstract}
  Visualizing the runtime behavior of programs can help programmers with
  targeted debugging and general understanding. For understanding
  complex programs, visualizations abstracted from the low-level code
  are most helpful, but this introduces new challenges: how does the
  programmer specify what to visualize, and how do we visualize complex
  data structures which aren't just primitive values? In this work, I
  present an approach to visualizing the behavior of user interfaces
  built with the Model-View-Update pattern. I present a prototype
  runtime visualization system built on the Redux library and argue
  that, by exploiting the natural abstraction characteristics of this
  application architecture, we can create useful runtime visualizations
  with minimal programmer effort.
\end{abstract}

@@

Much recent work in program visualization
\autocite{victora,guo2013,hoffswell2018a,pollock2019,kasibatla2018}
focuses on low-level details: showing the values of individual
variables, connected to individual lines of source code. This makes
sense for small programs, and for helping novices understand the basics
of programming languages. But these visualizations don't address the
needs of more experienced programmers working with larger programs.
Gaining a general understanding of a large program requires zooming out
far above the source code level.

This leads to the idea of \emph{abstract program visualization}:
creating abstract, program-specific views of runtime state or static
code, to help someone understand the program. This idea has been
explored in the context of teaching how algorithms work
\autocite{brown1984,stasko1990} and in understanding the behavior of
multithreaded Java programs \autocite{reiss2003,reiss2005}. In this
context, a tricky challenge emerges \autocite{reiss2007}: how can we
simultaneously tailor the visualization to the specific domain, while
enabling the programmer to create the visualization with minimal effort?
On the one hand, overly generic visualizations (as used in low-level
visualization systems like Python Tutor \autocite{guo2013}) will often
fail to capture the higher-level context of the specific program. On the
other hand, if a visualization is too specific and takes too much work
to create, it won't be reasonably possible for programmers to create the
visualization.

I think a promising strategy for approaching this problem is to create
runtime visualization systems coupled to a particular domain-specific
framework or language. Frameworks and DSLs often impose a particular
mental model, code architecture style, and other constraints that
usefully narrow the space of possible programs relative to a
general-purpose language. On the other hand, there are still many
different programs that can be built in one framework, so the effort of
building a visualization system can be amortized over thousands of
programs rather than concentrated on a single one.

One recent example of this strategy is visualizing runtime state in Vega
\autocite{hoffswell2018a}, a data visualization language. Because the
language is so domain-specific, the runtime visualization system is able
to make many contextual assumptions leading to a useful visualization.

In this work, I propose a runtime program visualization system for a
slightly more general category of applications: user interfaces built
with the Model-View-Update (MVU) architecture \autocite{fowler2020},
also commonly known as the Elm Architecture \autocite{czaplicki}. MVU
encourages the state of the interface to be centralized in a single data
structure, derived by running a pure reducer function over a stream of
events.

This architecture has been found to have many practical benefits for
program understanding and developer experience (e.g., making
``time-travel debugging'' relatively trivial to implement), but I think
it also has useful characteristics for abstract program visualization as
well. In particular, MVU naturally encourages programmers to define
abstractions that represent the essence of their application: 1) a
stream of semantically meaningful events, representing important changes
in the application, 2) a state object that represents all the core state
of the UI, organized in a hierarchical way. My hypothesis is that it's
possible to build a system that can visualize MVU interfaces with
relatively little additional effort from the programmer, because of
these natural characteristics of the architecture.

To begin to explore this idea, I've prototyped a runtime visualization
system on top of the popular Redux \autocite{zotero-621} Javascript
library. Within the limited scope of this project, I've focused
specifically on making a prototype specifically designed to visualize
the state of the TodoMVC demo application. I've designed some
visualization formats tailored to the state of that application, and
through my own usage I've begun to gain a preliminary understanding what
kinds of visualizations might be useful to programmers navigating the
state of MVU applications over time.

Much future work remains to fully flesh out this idea, including:

\begin{itemize}
\tightlist
\item
  developing a crisper understanding of the needs of programmers who are
  debugging MVU applications or trying to gain a general understanding
  of their behavior
\item
  generalizing the system so that it actually works with many Redux
  applications instead of just one, and has a way for a programmer to
  create a program-specific visualization
\end{itemize}

@@

Many systems have explored visualizing execution state at the level of
individual source lines, including Learnable Programming
\autocite{victora}, Python Tutor \autocite{guo2013}, Omnicode
\autocite{kang2017}, and Theia \autocite{pollock2019}.

@@

Other systems have explored fully abstract program visualization,
entirely disconnected from the source code. For example, Balsa
\autocite{brown1984} and Tango \autocite{stasko1990} show animated views
of algorithms operating, and Jive \autocite{reiss2003} and Jove
\autocite{reiss2005} visualize various high-level projections of the
execution of Java programs, e.g.~when different threads are running.

@@

I had some prior experience with the Redux Dev Tools system, which
provides the ability to inspect application state in Redux applications.
From this personal experience I identified two distinct use cases for a
runtime visualization:

\begin{itemize}
\tightlist
\item
  \textbf{Localizing within a trace}: \emph{Where do I need to rewind
  to, in order to inspect a particularly relevant point in an execution
  trace?} This is most often helpful when debugging a particular
  problem. Scrubbing back and forth while watching the UI change is a
  manageable solution, but it's inefficient, and sometimes the relevant
  state isn't directly visible in the UI.
\item
  \textbf{Generally understanding program behavior over time}:
  \emph{Overall, what happened as I interacted with the program?}
  Sometimes I'm not debugging a particular problem, and I'm more
  interested in just seeing general information about how a program is
  behaving over time. This is helpful when explaining the system's
  behavior to a new programmer who's preparing to work on the system, or
  for building up my own passive understanding of a system.
\end{itemize}

These two goals partially overlap, but can also lead in different design
directions. For example, localizing a specific point in a trace can
benefit from a more active interrogatory approach (e.g.~as explored in
the Whyline system \autocite{ko2004}), but general program understanding
might benefit from a more passive interaction style.

@@

\hypertarget{sec:design}{%
\section{Visualization Design}\label{sec:design}}

@@

In order to focus my effort on concretely understanding the utility of
visualizations, rather than building out infrastructure, I built a
visualization system for a specific application: the Redux
implementation of the TodoMVC GUI benchmark. TodoMVC is a basic todo
list UI where you can add, edit, delete, and complete todos. You can
filter the list of todos shown to either active or completed ones.

In Redux, the implementation stores an app state object as shown in
Figure~\ref{fig:todomvc-state}. There are Redux actions corresponding to
each of the main actions listed above, e.g.~``add todo'' and ``set
visibility filter''. Importantly, the Redux events capture an abstract,
semantically meaningful picture of the user's interactions: when adding
a new todo, the user's keystrokes are collected in the local state of a
React component, and only a single ``add todo'' event is triggered in
Redux once the user finally adds the new todo.

\begin{figure}
\hypertarget{fig:todomvc-state}{%
\centering
\includegraphics[width=1.82292in,height=2.08333in]{images/todomvc-state.png}
\caption{TodoMVC Redux application state}\label{fig:todomvc-state}
}
\end{figure}

@@

My initial idea, as shown in Figure~\ref{fig:mockup}, was to show the
current state of the application as a nested JSON tree, and then to show
small sparkline-style visualizations next to nodes of the tree. This
design draws some inspiration from \autocite{hoffswell2018a}, but
differs in that it uses visualization to annotate the application's
state tree, rather than its source code. The advantage of this design is
that it closely and directly links the current state to data from the
execution history, but that link also causes thorny problems---for
example, how do you deal with nodes that have disappeared from the
current state? Perhaps more concerningly, by tying the visualization to
the \emph{concrete} current state, it limits the ability of the
programmer to create a customized abstract view. Linking the
visualization to the Redux state is a much more abstract view than
linking to lines of source code, but it still presents a limitation.

As a result, in my next iteration I switched to a different layout,
shown in Figure~\ref{fig:timeline}: a vertically stacked list of small
visualizations of state over time. Each visualization can display an
arbitrary projection of the app's Redux state. Because the graphs are
horizontally aligned, it's easy to see how different aspects of the
app's state have changed in relation to each other. While I haven't
implemented this yet, I imagine that programmers would be able to
dynamically add visualizations to this list, specifying useful
projections of app state, and deciding what type of visualization to use
for each projection.

\begin{figure}
\hypertarget{fig:timeline}{%
\centering
\includegraphics[width=3.4375in,height=2.08333in]{images/timeline.png}
\caption{Timeline view of stacked visualizations}\label{fig:timeline}
}
\end{figure}

One thing lost in the timeline view is the concrete view of the app's
entire state. It's still useful to see this, so I added a separate panel
which displays that data. The user can scrub through history in the
timeline, ``pin'' the app at a particular point in time, and then use
the separate state view to drill into the app's concrete state at that
point.

@@

Here are the specific types of visualizations I prototyped (all shown in
Figure~\ref{fig:timeline}):

\emph{Action list}: I found that skimming a list of actions represented
as text (``ADD\_TODO'', ``EDIT\_TODO'') required a lot of conscious
reading effort. Instead, by choosing a colorful symbol for each action
in the app, we can take advantage of pre-attentive processing to more
quickly understand what actions have occurred in the execution trace. In
this case I chose symbols for all the actions in TodoMVC; more
generally, a programmer could specify a meaningful symbol for every
action in their application. In some cases it might be difficult to
choose meaningful and different symbols for all actions; falling back to
random symbols or colored dots could work as well. In using this tool
I've found that the symbolic action list makes it far easier to find a
point in an execution trace that I'm looking for.

@@

I implemented a working prototype version of this visualization system.
I built a plugin on top of the existing Redux Dev Tools, which provides
substantial infrastructure for inspecting and manipulating the state of
a Redux application.

I used the React and Redux frameworks to implement the main skeleton of
my system. The graphs are built in a combination of d3 and React---using
d3 for computing scales and positions, and React for SVG rendering.

>>>>

% Paper metadata (use plain text, for PDF inclusion and later
% re-using, if desired).  Use \emtpyauthor when submitting for review
% so you remain anonymous.
\def\plaintitle{Runtime Visualization for Model-View-Update GUIs}
\def\plainauthor{Geoffrey Litt}
\def\emptyauthor{}
\def\plainkeywords{Program visualization, program understanding, debugging}
\def\plaingeneralterms{Program visualization, program understanding, debugging}

@@

\begin{abstract}
  Visualizing the runtime behavior of programs can help programmers with
  targeted debugging and general understanding. For understanding
  complex programs, visualizations abstracted from the low-level code
  are most helpful, but this introduces new challenges: how does the
  programmer specify what to visualize, and how do they visualize
  complex data structures which aren't just primitive values?

  In this work, I present an approach to visualizing the behavior of
  user interfaces built with the Model-View-Update pattern. I present a
  prototype runtime visualization system built on the Redux library and
  argue that, by exploiting the natural abstraction characteristics of
  this application architecture, we can create useful runtime
  visualizations with minimal programmer effort.
\end{abstract}

@@

Much recent work in program visualization
\autocite{victora,guo2013,hoffswell2018a,pollock2019,kasibatla2018}
focuses on low-level details: showing the values of individual
variables, connected to individual lines of source code. This works well
for small programs, and for helping novices understand the basics of
programming. But these visualizations don't address the needs of more
experienced programmers working with larger programs. Gaining a general
understanding of a large program requires zooming out from individual
lines of code.

This leads to the idea of \emph{abstract program visualization}:
creating abstract, program-specific views of runtime state or static
code, to help someone debug or understand the program. This idea has
been explored in the context of teaching algorithms
\autocite{brown1984,stasko1990} and understanding the behavior of
multithreaded Java programs \autocite{reiss2003,reiss2005}. But abstract
visualizations create a new challenge \autocite{reiss2007}: how can we
enable the programmer to create program-specific abstract visualizations
with minimal effort? On the one hand, overly generic visualizations (as
used in most low-level visualization systems) will often fail to capture
the higher-level meaning of the specific program. On the other hand, if
a visualization takes too much work to create, it won't be realistic for
programmers to create the visualization in practice.

I think a promising strategy for approaching this problem is to create
runtime visualization systems coupled to a particular domain-specific
framework or DSL. Frameworks and DSLs occupy an intermediate place
between general-purpose languages and specific programs. They often
impose a particular mental model, code architecture style, and other
constraints that usefully narrow the space of possible programs relative
to a general-purpose language. On the other hand, there are still many
different programs that can be built in one framework, so the effort of
building a visualization system can be amortized over thousands of
programs rather than concentrated on a single one.

To concretely test this strategy, in this work I propose a runtime
program visualization system for user interfaces built with the
Model-View-Update (MVU) architecture \autocite{fowler2020}, also
commonly known as the Elm Architecture \autocite{czaplicki}. MVU
encourages the state of the interface to be centralized in a single data
structure, derived by running a pure reducer function over a stream of
events.

This architecture has many practical benefits for program understanding
and developer experience (e.g., automatically achieving time-travel
debugging), and I think it has useful characteristics for abstract
program visualization as well. In particular, MVU naturally encourages
programmers to define abstractions that represent the essence of their
application's behavior: 1) a stream of semantically meaningful events,
2) a state object that represents all the core state of the UI. My
hypothesis is that it is possible to visualize MVU interfaces with
relatively little additional effort from the programmer, because the
architecture has already required them to do much of the work of
abstracting.

I've prototyped a runtime visualization system on top of the popular
Redux \autocite{zotero-621} library, which implements MVU in Javascript.
Within the limited scope of this project, I've focused on making a
prototype specifically designed to visualize the state of the TodoMVC
demo application. I've designed some visualizations tailored to the
state of that application, and through my own usage I've begun to gain a
preliminary understanding what kinds of visualizations might be useful
to programmers navigating execution traces of MVU applications.

Much future work remains to fully flesh out this idea, including
developing a crisper understanding of the needs of programmers,
validating this system against those needs, and generalizing the system
so that it actually works with many Redux applications instead of just
one demo app.

@@

Many systems have explored visualizing execution state at the level of
individual source lines, including Learnable Programming
\autocite{victora}, Python Tutor \autocite{guo2013}, Omnicode
\autocite{kang2017}, Theia \autocite{pollock2019}, and Theseus
\autocite{lieber2014}.

@@

This work aims to provide a much more abstract view of the application's
behavior than any of these other projects, by avoiding doing any
visualization at the level of individual lines of code.

Other systems have explored this kind of abstract program visualization,
entirely disconnected from the source code. For example, Balsa
\autocite{brown1984} and Tango \autocite{stasko1990} show animated views
of algorithms operating, and Jive \autocite{reiss2003} and Jove
\autocite{reiss2005} visualize various high-level projections of the
execution of Java programs, e.g.~when different threads are running.

I'm not aware of much prior research on abstract program visualization
for user interfaces, although I still need to do a fuller literature
review. UI performance analysis tools or debuggers like the Redux Dev
Tools arguably fit into this category, but there aren't many tools that
employ data visualization techniques to display the internal state of
the application.

@@

I had some prior experience with the Redux Dev Tools debugger, which
provides the ability to inspect application state in Redux applications.
From this personal experience, I identified two distinct use cases for a
runtime visualization:

\begin{itemize}
\tightlist
\item
  \textbf{Localizing within a trace}: \emph{Where do I need to rewind
  to, in order to inspect a particularly relevant point in an execution
  trace?} This is most often helpful when debugging a particular
  problem. Scrubbing back and forth while watching the UI change is
  often workable, but it's inefficient. Also, sometimes the relevant
  state isn't directly visible in the UI, so I need to dig into a JSON
  object at each point in time to understand whether I've found the
  right point in the trace.
\item
  \textbf{Generally understanding program behavior over time}:
  \emph{Overall, what happened as I interacted with the program?}
  Sometimes I'm not debugging a particular problem, and I'm more
  interested in just seeing general information about how a program is
  behaving over time. For example, this is helpful when explaining the
  system's behavior to a new programmer who's preparing to work on the
  system, or when I'm trying to learn the basics of a codebase myself.
\end{itemize}

These two goals partially overlap, but can also lead in different design
directions. For example, localizing a specific point in a trace can
benefit from a more active interrogatory approach (e.g.~as explored in
the Whyline system \autocite{ko2004}), but general program understanding
might benefit from a more passive style, more akin to reading
documentation but accompanied by live demonstrations.

@@

@@

In order to focus my effort on concretely understanding the utility of
visualizations, rather than building out infrastructure, I built a
visualization system for a specific application: the TodoMVC GUI
benchmark. TodoMVC is a basic todo list UI where the user can add, edit,
delete, and complete todos. The user can also filter the list of todos
shown to either active or completed ones.

The Redux implementation of TodoMVC stores an app state object which
contains the list of todos, and the current state of the visibility
filter. There are actions corresponding to each of the main user
interactions listed above, e.g.~``add todo'' and ``set visibility
filter''. Importantly, the Redux events capture an abstract,
semantically meaningful picture of the user's interactions: when adding
a new todo, the user's keystrokes are collected in the local state of a
React component, and only a single ``add todo'' event is triggered in
Redux once the user finally adds the new todo.

@@

My initial idea, as shown in Figure~\ref{fig:mockup}, was to show the
current state of the application as a nested JSON tree, and then to show
small sparkline-style visualizations next to nodes of the tree. This
design draws some inspiration from \autocite{hoffswell2018a}, but
differs in that it uses visualization to annotate the application's
state tree, rather than its source code. The advantage of this design is
that it closely and directly links the current state to data from the
execution history, but that link also causes thorny problems---for
example, how do you deal with nodes that have disappeared from the
current state? Perhaps more concerningly, by tying the visualization to
the \emph{concrete} current state, it limits the ability of the
programmer to create a customized abstract view, removed from the
details of the state.

In my next iteration I switched to a different layout, shown in
Figure~\ref{fig:timeline}: a vertically stacked list of small
visualizations of state over time. Each visualization can display an
arbitrary projection of the app's Redux state. Because the graphs are
horizontally aligned, it's easy to see how different aspects of the
app's state have changed in relation to each other. While I haven't
implemented this yet, I imagine that programmers would be able to
dynamically add visualizations to this list, specifying useful
projections of app state, and deciding what type of visualization to use
for each projection.

\begin{figure}
\hypertarget{fig:timeline}{%
\centering
\includegraphics[width=3.4375in,height=1.77083in]{images/timeline.png}
\caption{Timeline view of stacked visualizations}\label{fig:timeline}
}
\end{figure}

One thing lost in the timeline view is the concrete view of the app's
entire state. It's still useful to see this, so I added a separate panel
which displays that data. The user can scrub through history in the
timeline, ``pin'' the app state at a particular point in time, and then
use the separate state view to drill into the app's concrete state at
that point.

@@

Here I describe the specific visualizations I prototyped for the
timeline view. These are shown in Figure~\ref{fig:timeline} from top to
bottom. (The video demo linked on the project page might be an easier
way to grasp the basics of each of these views)

\emph{Action list}: I found that skimming a list of actions represented
as text (\texttt{ADD\_TODO}, \texttt{EDIT\_TODO}, etc.) required a lot
of conscious reading effort. Instead, by choosing a colorful symbol for
each action in the app, we can take advantage of pre-attentive
processing to more quickly understand what actions have occurred in the
execution trace. In this case I chose symbols for all the actions in
TodoMVC; more generally, a programmer could specify a meaningful symbol
for every action in their application. In some cases it might be
difficult to choose meaningful and different symbols for all actions;
falling back to random symbols or colored dots could work as well. In
using this tool I've found that the symbolic action list makes it far
easier to find a point in an execution trace that I'm looking for.

\emph{Collection graph}: This visualization represents the contents of a
collection with a series of vertically stacked dots. The size of the
cluster of dots provides a rough sense of collection size, and the
programmer can more carefully examine the view to get an exact count.

Each dot has a color encoding for some attribute of the collection
element; in this case I've chosen to color the dots by whether the todo
is completed or not. Another available option is to color the dots by
identity---each unique element gets its own color.

I originally represented the list of todos with a line graph showing its
length, but this view allows us to display an additional dimension of
information for each todo. One corresponding weakness of this view is
that the size encoding doesn't offer too much information for
pre-attentive processing when there are more than a few elements so the
relative size change is small. It's not immediately obvious where in the
trace the number of todos changed, whereas a line graph makes it more
obvious. (One possible improvement would be to only show the dots on a
time step where the collection was changed.)

@@

I implemented a working prototype on top of the existing Redux Dev
Tools, which provides substantial infrastructure for inspecting and
manipulating the state of a Redux application. My tool is implemented as
a Redux Dev Tools ``monitor'' which can plug in to those existing APIs.

I used the React and Redux frameworks to implement the main skeleton of
my system. The graphs are built in a combination of d3 and React. I use
d3 for computing scales and positions, and then React for actually
rendering out SVGs.
